---
title: 'Lab #3'
author: "Ashley Sackpraseuth"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part one: Extracting and loading data

```{r}
library(tidyverse)

dat <- readxl::read_xls("GSS.xls") %>%
  rename(year = `Gss year for this respondent`, id = `Respondent id number`, marital = `Marital status`, age = `Age of respondent`, educ = `Highest year of school completed`, sex = `Respondents sex`, party = `Political party affiliation`, happy = `General happiness`, ballot = `Ballot used for interview`)
dat$age <- as.numeric(dat$age)
dat$educ <- as.numeric(dat$educ)
dat$party <- factor(dat$party, levels=c("Strong democrat", "Not str democrat", "Ind,near dem", "Independent", "Ind,near rep", "Not str republican", "Strong republican", "Other party", "No answer", "Don't know"))
dat$happy <- factor(dat$happy, levels=c("Not applicable", "Don't know", "No answer", "Not too happy", "Pretty happy", "Very happy"))
dat$ballot <- factor(dat$ballot, levels=c("Ballot a", "Ballot b", "Ballot c", "Ballot d", "Not applicable"))
dat$marital <- factor(dat$marital, levels=c("No answer", "Never married", "Married", "Separated", "Divorced", "Widowed"))
dat$sex <- factor(dat$sex)
```

## Part two: Exploration 

1. Find the average years for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.

    ```{r}
    pol_party_group <- dat %>%
                            group_by(party) %>% 
                            summarize(avg_educ = mean(educ, na.rm=TRUE),
                            nrows = n()
                        )
    
    pol_party_group[2] <- round(pol_party_group[2], 0)
    pol_party_group <- pol_party_group[order(-(pol_party_group$avg_educ)),]
    print(pol_party_group)
    ```

    Based on this table, the respondents who have more years of education (13 years) tend to be more independent or associate with other party or republican. "Democrat" and "Independent" tend to have slightly less years of education (12). "Don't know" and "No answer" tend to include respondents with the least years of education (11-12). 

2. Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has strictly more years of education than the median in the same party affiliation. Print the structure of the data frame.

    ```{r}
    educ_pol_party <- dat %>%
                            group_by(party) %>% 
                            mutate(med_educ = median(educ, na.rm=TRUE)) %>%
                            filter(age >= med_educ)
    
    educ_pol_party[10] <- round(educ_pol_party[10], 0)
    print(str(educ_pol_party))
    ```

3. How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.

    ```{r}
      
    # Making a new happy column based on the prompt 
      dat$happy_redefined = NULL
    
      happy <- unique(dat$happy)
      happy
      
      happy_marriage <- dat %>%
                            mutate(happy_redefined = factor(case_when(happy == 'Not too happy' ~ "Not happy",
                                                               happy == 'Pretty happy' ~ "Happy",
                                                               happy == 'Very happy' ~ "Happy",
                                                               happy == 'No answer' ~ "No answer",
                                                               happy == 'Very happy' ~ "Happy",
                                                               happy == "Don't know" ~ "Don't know",
                                                               happy == 'Not applicable' ~ "Not-applicable",
                                                               ), levels=c("Happy", "Not happy", "Not-applicable", "Don't know", "No answer"))
                                 )
    
      
    # Numerical summary 
    happy_table <-table(happy_marriage$marital, happy_marriage$happy_redefined)
    print(happy_table)
    
    data_sums <- rowSums(happy_table)
    data_sums
        
        
    # Visualizations 
    ggplot(happy_marriage, aes(x=marital, fill=happy_redefined)) + geom_bar() + ggtitle("Happiness of Marital Statuses") + xlab("Marital status") + ylab("Count") 
    
      
      
    ggplot(happy_marriage, aes(x=marital, fill=happy_redefined)) + geom_bar(position = 'fill') + ggtitle("Happiness of Marital Statuses Proportion") + xlab("Marital status") + ylab("Count") 
    ```

    According to both the table and the first plot, majority of the respondents are married. To ensure we are getting an accurate view of how marital status affects happiness, the second plot was made. The second plot represents the percentages of the populations being measured rather than the counts themselves. According to the second plot, those who are married are the happiest, followed by those who were never married, divorced, widowed, separated, and then 'no answer'. While the exact reasons for this cannot be concluded from this chart alone, one possible explanation can be that those who are married are the happiest because they have someone to share their happiness with. One possible explanation for those who are never married ranking second on the happiness scale is they know they are not bound to anything; if they are unhappy with their relationship they can leave without major consequences. Divorced, separated, and widowed rank lowest on the happiness scale because they have endured financial and emotional hardships.

4. Does party affiliation affect the relationship you found in the last question?

    ```{r}
    hap_mar_pol <- happy_marriage %>% 
                    group_by(happy_redefined, marital, party)
    
    
    
    hap_table <- table(happy_marriage$happy_redefined, happy_marriage$marital, happy_marriage$party)
    
        hap_table <- data.frame(hap_table)
        names(hap_table)[1] <- 'Happiness' 
        names(hap_table)[2] <- 'Marital status'
        names(hap_table)[3] <- 'Politcal party'
        names(hap_table)[4] <- 'Frequency'
        
     hap_table <- hap_table[order(-(hap_table$Frequency)),]   
    
    print(head(hap_table))
    
    
    #Visualization
    ggplot(happy_marriage, aes(x=marital, fill=happy_redefined)) + geom_bar(stat="count")+ facet_wrap(~party) + theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust = 1))
    ```

    Political party affiliation does not appear to affect the relationship I found in the last question.

5. How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?

    ```{r}
    # Numerical summary
      dat$`Highest year of school completed` <- as.numeric(dat$educ)
    
    
      mar_educ <- dat %>%
                        group_by(marital) %>% 
                        summarize(avg_educ = mean(educ, na.rm=TRUE)
                        )
      mar_educ <- mar_educ[order(-(mar_educ$avg_educ)),]
      print(mar_educ)
    
    # Visualization
      ggplot(dat, aes(marital,educ, fill=marital)) + geom_boxplot() + ggtitle("Education level of Marital Statuses")
    ```

    Based on both the table and plot, those who received a shorter education were more likey to separate or become widowed. Surprisingly, those who chose not to answer the marital status question had the highest education, followed by those who were never married, then those who were divorced. Those who are married fall in the middle. One possible explanaation for "No answer" and "Never married" groups having the highest level of education is they value education/career-building over personal relationships, which is very common in this day and age. 



6. Explore two more interesting questions, and answer those questions using the GSS data.

    One question to explore is how does general happiness compare to age? Another question is how does highest number of years completed school compare to general happiness? 

    To answer the first question 
    ```{r}
    # Numerical summary
    avg_respondent <- mean(dat$age, na.rm = TRUE)
    avg_respondent
    
    happy_age <- dat %>%
                            group_by(happy) %>% 
                            summarize(avg_happy_age = mean(age, na.rm=TRUE),
                            nrows = n()
                        )
    
    colnames(happy_age)
    happy_age[2] <- round(happy_age[2], 0)
    ref <- c("Very happy", "Pretty happy", "Not too happy", "Don't know","No answer", "Not applicable")
    ref
    happy_age <- factor(happy_age$avg_happy_age, levels = ref)
    print(happy_age)
    
    
    #Visualization
    ggplot(dat, aes(x=happy,y=age, fill = happy)) + geom_boxplot() + ggtitle("Age vs. General Happiness")
    ```

    The average age of all respondents of this survey is 45 years old. Based on the table and plot, age of respondent does not seem to affect happiness. The only notable trend that can be seen is that at age 55, most respondents do not not know if they are happy.

    Another question to explore is how to does general years of school completed compare to general happiness? 
    ```{r}
    happy_educ <- dat %>%
                            group_by(happy) %>% 
                            summarize(avg_educ = mean(educ, na.rm=TRUE),
                            nrows = n()
                        )
    
    colnames(happy_educ)
    happy_educ[2] <- round(happy_educ[2], 0)
    print(happy_educ)
    
    
    #Visualization
    ggplot(dat, aes(x=happy,y=educ, fill = happy)) + geom_boxplot() + ggtitle("Age vs. General Happiness")
    ```

    Based on the plot and table, it looks like education level does not affect happiness. This is surprising to me because I expected those with a higher education to be happier because usually a higher education is associated with a higher paying job.