---
title: "Lab 3 - Team"
author: "Brett White and Ashley Sackpraseuth"
output: html_document
---

## Part one: Extracting and loading data

The code chunk below loads the data and cleans up the data frame.

```{r load}
library(tidyverse)
dat <- readxl::read_xls('GSS.xls') %>%
  rename(year = `Gss year for this respondent`, id = `Respondent id number`, marital = `Marital status`, age = `Age of respondent`, educ = `Highest year of school completed`, sex = `Respondents sex`, party = `Political party affiliation`, happy = `General happiness`, ballot = `Ballot used for interview`)
dat$age <- as.numeric(dat$age)
dat$educ <- as.numeric(dat$educ)
dat$party <- factor(dat$party, levels=c("Strong democrat", "Not str democrat", "Ind,near dem", "Independent", "Ind,near rep", "Not str republican", "Strong republican", "Other party", "No answer", "Don't know"))
dat$happy <- factor(dat$happy, levels=c("Not applicable", "Don't know", "No answer", "Not too happy", "Pretty happy", "Very happy"))
dat$ballot <- factor(dat$ballot, levels=c("Ballot a", "Ballot b", "Ballot c", "Ballot d", "Not applicable"))
dat$marital <- factor(dat$marital, levels=c("No answer", "Never married", "Married", "Separated", "Divorced", "Widowed"))
dat$sex <- factor(dat$sex)
```

## Part two: Exploration 

1. Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.

    ```{r, dependson = "load"}
    dat %>%
      group_by(party) %>%
      summarize(mean_educ = mean(educ, na.rm=TRUE))
    ```

    Republicans tend to have the highest number of years of education followed by Democrats, then Independent or other parties.

2. Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has strictly more years of education than the median in the same party affiliation. Print the structure of the data frame.

    ```{r, dependson = "load"}
    better_educ <- dat %>%
      group_by(party) %>%
      filter(educ > median(educ, na.rm=TRUE))
    str(better_educ)
    ```

3. How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.

    ```{r, dependson = "load"}
    ggplot(dat %>% filter(!happy %in% c("Not applicable", "Don't know", "No answer") & marital != "No answer"), aes(happy)) + geom_bar() + facet_wrap(~marital, scales = "free_y") + theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust = 1))+ ggtitle("General Happiness vs. Marital Status") + xlab("General happiness") + ylab("Count")
    dat %>%
      filter(!happy %in% c("Not applicable", "Don't know", "No answer") & marital != "No answer") %>%
      group_by(marital) %>%
      summarize(perc_happy = sum(happy %in% c("Pretty happy", "Very happy")) / n())
    ```

    Of the 5 martial status groups (never married, married, separated, divorced, and widowed), married individuals reported to be the happiest with `r dat %>% filter(happy %in% c("Pretty happy", "Very happy") & marital == "Married") %>% nrow` out of `r dat %>% filter(marital == "Married") %>% nrow` of these individuals indicating that they were pretty happy or very happy. Those who never married follow with `r dat %>% filter(happy %in% c("Pretty happy", "Very happy") & marital == "Never married") %>% nrow` out of `r dat %>% filter(!happy %in% c("Not applicable", "Don't know", "No answer") & marital == "Never married") %>% nrow` (`r round((dat %>% filter(happy %in% c("Pretty happy", "Very happy") & marital == "Never married") %>% nrow) / (dat %>% filter(!happy %in% c("Not applicable", "Don't know", "No answer") & marital == "Never married") %>% nrow) * 100, digits=2)`%) of these individuals indicating that they were pretty happy or very happy. Divorced and widowed individuals follow with approximately 80% of those individuals reporting they were pretty happy. Separated individuals had the lowest percentage of people reporting they were happy with `r dat %>% filter(happy %in% c("Pretty happy", "Very happy") & marital == "Separated") %>% nrow` out of `r dat %>% filter(marital == "Separated") %>% nrow` of these individuals indicating that they were pretty happy or very happy. Many of these individuals are pretty happy but not too happy. We used the general happiness and marital status variables to arrive at these findings.

4. Does party affiliation affect the relationship you found in the last question?

    ```{r, dependson = "load"}
    ggplot(dat %>% filter(!happy %in% c("Not applicable", "Don't know", "No answer") & marital != "No answer" & !party %in% c("No answer", "Don't know")), aes(happy)) + geom_bar() + facet_grid(marital~party, scales = "free_y") + theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust = 1)) + ggtitle("General Happiness of Political Party Affiliation vs. Marital Status") + ylab("Count") + xlab("General Happiness")
    dat %>%
      filter(!happy %in% c("Not applicable", "Don't know", "No answer") & marital != "No answer" & !party %in% c("No answer", "Don't know")) %>%
      group_by(marital, party) %>%
      summarize(perc_happy = sum(happy %in% c("Pretty happy", "Very happy")) / n())
    ```    

    Married individuals who were strong republicans were found to be the happiest with 95% of them reporting either being very happy or pretty happy. All other political associations for married individuals had mostly pretty happy reports. The other martial statuses and political affiliations showed no changes in distributions. We used the general happiness, marital status, and political party affilication variables to arrive at these findings.

5. How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?

    ```{r, dependson = "load"}
    dat %>%
      group_by(marital) %>%
      summarize(mean_educ = mean(educ, na.rm=TRUE))
    ```

    ```{r, dependson = "load"}
    ggplot(dat %>% filter(!is.na(educ) & marital != "No answer"), aes(educ)) + geom_histogram(binwidth=1) + facet_wrap(~marital, scales = "free_y") + ggtitle("Highest Year of School Completed vs. Marital Status") + ylab("Count") + xlab("Number of years completed in school")
    ```

    The histograms above appear to show that regardless of marital status, the distribution of number of years completed in school is relatively the same. Widowed and separated have the widest range of number of years completed in school. Based on the table, those who chose not to answer completed the highest years of education with an average of 15.5 years, followed by never married, divorced, married, separated, then widowed. We used the years of education and marital status variables to arrive at these findings.

6. Explore two more interesting questions, and answer those questions using the GSS data.

    ```{r, dependson = "load"}
    ggplot(dat %>% filter(!happy %in% c("Not applicable", "No answer", "Don't know") & !party %in% c("No answer", "Don't know")), aes(happy)) + geom_bar() + facet_wrap(~party, scales = "free_y") + theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust = 1)) + ggtitle("General Happiness vs. Political Party Affiliations") + xlab("General Happiness") + ylab("Count")
    dat %>%
      filter(!happy %in% c("Not applicable", "No answer", "Don't know") & !party %in% c("No answer", "Don't know")) %>%
      group_by(party) %>%
      summarize(perc_happy = sum(happy %in% c("Pretty happy", "Very happy")) / n())
    ```

    From the barplots and table above it can be seen that republicans tend to be happier than democrats, other parties, and independents. More than 90% of individuals affiliated with the republican party reported being pretty happy or very happy while only approximately 85% of democrats, other parties, and independents reported being pretty happy or very happy. We used the general happiness and political party affilication variables to arrive at these findings.

    ```{r, dependson = "load"}
    ggplot(dat %>% filter(!happy %in% c("Not applicable", "No answer", "Don't know") & marital != "No answer"), aes(happy)) + geom_bar() + facet_grid(sex~marital) + theme(axis.text.x = element_text(angle = 90, vjust = 0.25, hjust = 1)) + ggtitle("Gender vs. Marital Status Happiness Levels") + ylab("Count") + xlab("General Happiness")
    dat %>%
      filter(!happy %in% c("Not applicable", "No answer", "Don't know") & marital != "No answer") %>%
      group_by(marital, sex) %>%
      summarize(perc_happy = sum(happy %in% c("Pretty happy", "Very happy")) / n())
    ```

    Based on the barplots and table above, it can be seen that there is generally no difference in happiness between genders for each marital status. Females of any marital status tend to be slightly happier than males, and divorced and widowed females tend to be significantly happier than divorced and widowed males. We did discover that there were more divorced and widowed females from whom data was collected than males. We used the general happiness, marital status, and gender variables to arrive at these findings.